Machine Learning for Predictive Auto-Tuning with Boosted Regression Trees

### 2012-03-11 paper/final.tex
final touchup
final spellcheck

\author{
\alignauthor James Bergstra\\
\affaddr{\small{Rowland Institute at Harvard}}\\
\affaddr{\small{Cambridge MA, US}}\\
\email{\emph{\small{bergstra@rowland.harvard.edu}}}
% 2nd author
\alignauthor Nicolas Pinto\\
\affaddr{\small{Rowland Institute at Harvard}}\\
\affaddr{\small{McGovern Institute at MIT}}\\
\affaddr{\small{Cambridge MA, USA}}\\
\email{\emph{\small{pinto@rowland.harvard.edu}}}
% 3rd author
\alignauthor David Cox\\
\affaddr{\small{Rowland Institute at Harvard}}\\
\affaddr{\small{Cambridge MA, USA}}\\
\email{\emph{\small{cox@harvard.rowland.edu}}}
}

@@

Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
\citep{schapire:2001,friedman:2002}
In a recent empirical study of a range of machine learning regression problems,
boosted decision trees were found to be among the best and easiest models to
apply~\citep{caruana+niculescu-mizil:2006}.
On each boosting iteration, a regression tree is fit to the residual error remaining
after all previously-fit models have made their predictions.
There are essentially three parameters that control the boosted regression tree
training procedure:
1) the depth of tree constructed on each boosting iteration,
2) the minimum number of examples to allow at a regression tree leaf, and
3) the number of trees constructed by boosting.
We did not attempt a systematic study of the effect of these variables on performance.
We chose a maximum depth of 4, so that each new tree would be a {\em weak}
learner without too much capacity;
a minimum number of examples per leaf of 10,
so that the ensemble would resist overfitting;
and 100 iterations of boosting, after which the residual error on training
data appeared to have reached a minimum.  Test examples were not used to
choose these values.

@@

Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is
in this setting.  Taking the GTX 580 as an example, and averaging across the
range of problem configurations in our study, we find that empirically
auto-tuned implementations are on average about 50\% faster than the reference
implementation.  The reference in turn, is about 50\% faster than
implementations that were empirically auto-tuned for a \emph{randomly chosen}
different argument configuration.  This shows that it is genereally not enough
to auto-tune for particular argument configurations; instead it is important to
choose the right kernel for the job for each unique argument configuration
(input-dependent auto-tuning).

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580_R2.pdf} % on vader
\caption{The effect of the invalid configuration score ($\zeta$) and training set
size on predictive auto-tuning.
Models trained on as few as 25 configurations outperformed the reference.
Models trained on 50 or more configurations rivaled empirical auto-tuning.
Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
}
\label{fig:fig_ntrain}
\end{figure}

@@

% XXX: how to make sense of these paramters without code listing?

@@

% we believe that our approach is the most practial one to get high perf on

@@

% auto-tuning required hardward expertise *and* application-domain knowledge, we

@@

% more features (this is the hardest, and most manual part, may require
% domain-specific knowledge, but not necessariyly, e.g. microbenchmarking)
% * features derived from micro-benchmarking \citep{wong2010demystifying}
% * features derived from performance limiter analysis \citep{micikevicius2010analysis}

@@

% * more instrumentation
% * more data-layout
% * more threading strategies to lower latency
% (finding a good representation is key, again...)
%
% online distributed auto-tuning with learning, especially w/ HT-based
% experiments \cite{HT} our method is indeed very well suited for parallel
% explorating of multiple configurations for learning (in a cluster, e.g. in HT
% context) this may get us \emph{more} than linear speed ups for better accuracy,
% especially when more features will be present (see above)
%
% interpret the learned model and confront the performance against theoreticial
% (or approximated) models of mem/com/compute, arithmetic intensity, etc. to
% understand performance limiters, the hardware, etc.

>>>>

\author{
\alignauthor James Bergstra\\
\affaddr{\small{Rowland Institute at Harvard}}\\
\affaddr{\small{Harvard University}}\\
\affaddr{\small{Cambridge MA, USA}}\\
\email{\emph{\small{bergstra@rowland.harvard.edu}}}
% 2nd author
\alignauthor Nicolas Pinto\\
\affaddr{\small{Rowland Institute at Harvard}}\\
\affaddr{\small{McGovern Institute at MIT}}\\
\affaddr{\small{Cambridge MA, USA}}\\
\email{\emph{\small{pinto@rowland.harvard.edu}}}
% 3rd author
\alignauthor David Cox\\
\affaddr{\small{Rowland Institute at Harvard}}\\
\affaddr{\small{Harvard University}}\\
\affaddr{\small{Cambridge MA, USA}}\\
\email{\emph{\small{cox@harvard.rowland.edu}}}
}

@@

Boosting is an iterative procedure for constructing an {\em ensemble} of regression trees that is coordinated to fit training examples as accurately as possible.
\citep{schapire:2001,friedman:2002}
In a recent empirical study of a range of machine learning regression problems,
boosted decision trees were found to be among the best and easiest models to
apply~\citep{caruana+niculescu-mizil:2006}.
On each boosting iteration, a regression tree is fit to the residual error remaining
after all previously-fit models have made their predictions.
There are essentially three parameters that control the boosted regression tree
training procedure:
1) the depth of tree constructed on each boosting iteration,
2) the minimum number of examples to allow at a regression tree leaf, and
3) the number of trees constructed by boosting.
We did not attempt a systematic study of the effect of these variables on performance.
We chose a maximum depth of 4, so that each new tree would be a {\em weak}
learner without too much capacity;
a minimum number of examples per leaf of 10,
so that the ensemble would resist over-fitting;
and 100 iterations of boosting, after which the residual error on training
data appeared to have reached a minimum.  Test examples were not used to
choose these values.

@@

Figure~\ref{fig:allstars} shows the effectiveness of empirical auto-tuning is
in this setting.  Taking the GTX 580 as an example, and averaging across the
range of problem configurations in our study, we find that empirically
auto-tuned implementations are on average about 50\% faster than the reference
implementation.  The reference in turn, is about 50\% faster than
implementations that were empirically auto-tuned for a \emph{randomly chosen}
different argument configuration.  This shows that it is generally not enough
to auto-tune for particular argument configurations; instead it is important to
choose the right kernel for the job for each unique argument configuration
(input-dependent auto-tuning).

@@

\begin{figure}[!t]
\centering
\includegraphics[width=\linewidth]{illustrator/fig_ntrain_munctional0_580_R2.pdf} % on vader
\caption{The effect of the invalid configuration score ($\zeta$) and training set
size on predictive auto-tuning.
Models trained on as few as 25 configurations outperformed the reference.
Models trained on 50 or more configurations rivalled empirical auto-tuning.
Moderate values of $\zeta$ between $\log(0.25)$ and $\log(0.75)$ are best.
}
\label{fig:fig_ntrain}
\end{figure}

@@

% XXX: how to make sense of these parameters without code listing?

@@

% we believe that our approach is the most practical one to get high perf on

@@

% auto-tuning required hardware expertise *and* application-domain knowledge, we

@@

% more features (this is the hardest, and most manual part, may require
% domain-specific knowledge, but not necessarily, e.g. microbenchmarking)
% * features derived from micro-benchmarking \citep{wong2010demystifying}
% * features derived from performance limiter analysis \citep{micikevicius2010analysis}

@@

% * more instrumentation
% * more data-layout
% * more threading strategies to lower latency
% (finding a good representation is key, again...)
%
% online distributed auto-tuning with learning, especially w/ HT-based
% experiments \cite{HT} our method is indeed very well suited for parallel
% exploration of multiple configurations for learning (in a cluster, e.g. in HT
% context) this may get us \emph{more} than linear speed ups for better accuracy,
% especially when more features will be present (see above)
%
% interpret the learned model and confront the performance against theoretical
% (or approximated) models of mem/com/compute, arithmetic intensity, etc. to
% understand performance limiters, the hardware, etc.
